Different forms of uncertainty differently affect the evolution of social learning

### 2022-07-19 ms.tex
content filled in mostly; needs more specificity and more revisions

\textbf{Effective lifespan.}
\mt{Restructure: lead with ``It stands to reason'' sentence or similar that gets
to the point of this par more quickly, that lifespan affects uncertainty.}
Learning can be viewed as a process by which an individual reduces or otherwise manages their uncertainty \cite{jacobs2011bayesian,clark2013whatever}. It is an iterative process by which one repeatedly acquires information, forms representations and predictions, and tests and refines those representations and predictions. It therefore stands to reason that the more opportunities one has for learning, the more one can in principle reduce one's uncertainty, assuming the environment has not changed in the meantime. Similarly, a reduction in the number of opportunities to learn will increase the uncertainty about whether one has accurately learned about the available behavioral options and their associated payoffs. We refer to the number of learning opportunities as an individual's effective lifespan to imply that it is the number of opportunities to learn throughout the individual's lifespan that matters here, not necessarily how long they live in the absence of relevant learning opportunities.

@@

We operationalized the four ``principal components'' of uncertainty that affect the
ability of agents to find and choose the optimal behavior. Recall these varieties of
uncertainty are \emph{environmental variability}; \emph{payoff ambiguity};
\emph{selection set size}; and the \emph{effective lifespan}. Environmental
variability is the probability that the behavior yielding the optimal payoff changes
from one generation to the next. We do not consider cases where the optimal behavior
varies within a generation. This is empirically justified since cultural evolution
occurs when meta-cognitive strategies and information are transferred or not via
learning and social influence, instead of truly by physical reproduction and death
as in genetic evolution.  The selection set size is the total number of possible
behaviors agents choose from, assumed to be constant within a simulation.  We
operationalized payoff ambiguity by manipulating the payoff difference between optimal
and non-optimal behaviors. The effective life span is the
number of time steps in which agents perform behaviors within a generation. 
Our model is not spatial, so agents have free, instant
access to all behaviors, and there is no spatial uncertainty, 
as might be the case for humans in a computer-based lab
experiment~\cite{McElreath2005,Morgan2012}.  Spatial dimensions could also be
considered irrelevant if costs associated with accessing a behavior were equal,
e.g., if it was assumed to be equally easy for bumblebees to fly to alternative
flower ``species''~\cite{Baracchi2018} or for great tits to fly to different food
sources~\cite{Aplin2017}.  Each simulated environment provides a distinct challenge
to the agents, which will lead agents to differently evolve to become social
learners or not depending on which of those two strategies optimizes net payoffs.

\emph{Agents} here are autonomous, intelligent, simulated problem solvers. Each
agent evaluates its environment and chooses how to act within that environment based
on a set of empirically-motivated rules.  Our agents select which behavior to do
probabilistically, weighted by the softmax function applied to the mean payoff they
have observed for each behavior. The softmax function is a biologically plausible
generalization of behavior selection under uncertainty that enables agents to often
greedily exploit the most lucrative behavior (based on their observations) but also
to sometimes explore alternatives~\cite{Schulz2019,Collins2013,Daw2006,Yechiam2005}.
Such individual-level intelligence is a critical adaptation that has apparently
co-evolved with the capacity for social learning, and has a marked influence on the
cultural evolution of social learning as we will demonstrate in our Analysis.

@@

The model environment provides behaviors for agents to perform. The structure
of the environment is specified by setting several parameters, including
three of the uncertainty parameters: environmental variability, selection set size, 
and payoff ambiguity. 
Each behavior is indexed by $b$ and modeled as a ``one-armed bandit'', which yields 
Bernoulli-distributed payoffs of 1 with probability $\pi_b$ and 0 with 
probability $1 - \pi_b$. $\pi_b$ is equivalently the expected
payoff of behavior/bandit $b$. The model environment affords $B$ behaviors, where
$B$ is the \emph{selection set size} uncertainty parameter, which
is constant throughout the course of a given simulation.

The model environment is dynamic: which behavior is optimal changes between
generations with probability equal to the \emph{environmental variability}, $u$.
At the beginning of each generation, one bandit is optimal and yields expected
payoff $\pihigh$; the other $B-1$ behaviors yield a lower payoff, $\pilow$. 
The \emph{payoff ambiguity} here between optimal and non-optimal behavior
payoffs is defined by $\pihigh - \pilow$. We set $\pihigh = 0.9$ to be constant
across all simulations, so our analysis only varies payoff ambiguity by varying
$\pilow$. Specifically, payoff ambiguity increases with $\pilow$. 
By systematically varying these uncertainty parameters we change
the nature of the challenge faced by the agents. Each parameter individually 
can increase or decrease ``uncertainty,'' but it is not at all clear \emph{a priori}
how these uncertainties combine to affect the evolution of social learning.
These uncertainty parameters and the values we tested in our simulations
are summarized in Table~\ref{tab:uncertaintyParameters}.

@@

In this model, being a social learner is beneficial when it is more likely
on average to provide greater net payoffs than being an individual learner,
since evolution here acts as an optimization routine. 
Note that, unlike some models in the literature, it is not catastrophic for an 
individual to receive outdated information\cite[, e.g]{Rogers1988}. Instead,
agents can explore alternatives and greedily perform the observed optimal behavior
if they learned outdated information due to environmental variability. We rely
on this mechanism in the Analysis to explain patterns we observe where
social learning evolves when $u > 0.5$, meaning that the
optimal behavior will switch between generations more often than not. If 
agents could not adjust to changing conditions like this, it seems like 
social learning would not evolve when $u > 0.5$.

@@

\paragraph{Intergenerational dynamics}
In between generations agents first reproduce via asexual haploid reproduction; 
then new child agents learn from the
previous generation if they are social learners; and finally all agents from the
previous generation then die off (Figure~\ref{fig:schematic}C). 
$N$ reproducers are selected with replacement over $N$ independent draws, 
biased by performance:
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce in one draw}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
A child inherits its parent's social learning trait $s_i$ without mutation.
A social learner child with $s_i = 1$ learns from a teacher from its parent's
generation, including possibly their parent (oblique learning). 
A child selects its teacher by first selecting $N_T$ prospective
teachers from the population completely at random, then selecting the one with
highest net payoffs $\pi_i$ as their teacher, with ties broken randomly. While it
is somewhat arbitrary to first subset $N_T$, this is a conservative choice that
represnts the fact that access to the entire population is not generally guaranteed.
Social learners adopt their chosen teacher's ledger of expected payoffs $\ledger$,
and set all $c_{ib} = 1$ so expected payoffs remain within $[0, 1]$.  Asocial
learners ($s_i = 0$) are initialized with all $c_{ib} = 0$ and $\ledger = 0$ 
(Figure~\ref{fig:schematic}C). The
newly initialized population then begins behavior selection and payoff accumulation
for another $L$ time steps.

After reproduction and die-off, the optimal behavior changes with probability
$u$. If a random number draw indicates the behavior should change, the new
behavior is chosen from all behaviors but the current optimal behavior.

@@

Payoff ambiguity had a modest effect when $\pilow = 0.1$ increased to $\pilow = 0.45$,
causing social learning to be favored across more values of $u$, across all
selection set sizes and lifespans (first two rows of Figure~\ref{fig:mainResults}). 
This indicates that increased payoff ambiguity significantly increases the 
difficulty of discerning which behavior is optimal since agents evolve to 
risk social learning at greater values of $u$.
When $\pilow = 0.8$ the largest effect is smoothing and flattening of the $\meansl$ curves,
which indicates more drift (bottom row of Figure~\ref{fig:mainResults}). 
Initially this results in less frequent evolution
of social learning for the same values of $u$ and $B$. But after 
the $u$ value where $\meansl \approx 0.5$, social learning evolves more frequently than
for the same $u$ and $B$. When $\pilow \to \pihigh$, agents do better and better
whether they are social learners or not. We see that this leads to less penalty
for trying social information before possibly having to correct course due to
outdated social information.

@@

@@

By systematizing four common uncertainty dimensions we developed a model of
the evolution of social learning that
predicts how different forms of uncertainty might interact to foster or suppress
social learning.  The model reproduced classic predictions that social learning can
yield higher payoffs than individual learning when uncertainty is sufficiently
limited.  By disambiguating and organizing various operationalizations of
uncertainty, we developed a more nuanced theoretical understanding of which forms
of uncertainty interact in which ways to make social learning more or less
beneficial compared to asocial learning.
Our model did not make any \emph{a priori} assumptions about whether it was
more costly for agents to be social or asocial learners, as some studies have
assumed or imposed~\cite{Rogers1988,Feldman1996,Morgan2012}. 
Instead, the uncertainty structure determined whether
or not social learning was beneficial or not, and evolution directed successive
generations towards the optimal strategy.
Such detailed theoretical models of uncertainty
are critical for understanding how humans adapt to an uncertain and
rapidly changing world, both to mitigate existential
threats~\cite{Moya2020,Jones2021}, and to capitalize on new opportunities such as
transitioning to clean energy use and
production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.

Our model necessarily made some simplifying assumptions that, while
empirically justified in many cases, may not always hold.
\mt{Review the ones mentioned in model section: non-spatial, no within-generation
environmental variability; no horizontal learning}.
Furthermore, we only considered
success-biased learning, although conformity is theoretically important as
well~\cite{Muthukrishna2016a,Smaldino2018b}. Another implicit assumption
in our model was that the number of behaviors was constant, as was the relationship
between their payoffs---one was always optimal, and the rest were equally
suboptimal. However, this fails to account for evolutionary feedback which creates
new behavioral affordances as time progresses, e.g.\ via niche 
construction~\cite{Smaldino2012,Heras-Escribano2020}.

@@

Social learning is an immensely complex phenomenon. In just our simple model, we
found that its evolution is similarly complex, dependent on several interrelated
factors. By systematically identifying common theoretical assumptions about
uncertainty in the environments and contexts in which social learning evolves,
we were able to develop a more systematic understanding of which uncertainty
factors. This work informs most directly our current understanding of the
evolution of social learning, but
has broader importance for the behavioral science of problem solving in uncertain
conditions, and for the design of artificial social intelligent systems.
Our model and its software implementation were designed to be modular and 
therefore extensible.  Indeed, we plan to extend this model to further develop 
a more detailed theoretical understanding of the evolution of social learning, 
and we hope others do, too.

>>>>

\textbf{Effective lifespan.}
\mt{Restructure: lead with ``It stands to reason'' sentence or similar that gets
to the point of this par more quickly, that lifespan affects uncertainty. Also this Effective
Lifespan section seems too long.}
Learning can be viewed as a process by which an individual reduces or otherwise manages their uncertainty \cite{jacobs2011bayesian,clark2013whatever}. It is an iterative process by which one repeatedly acquires information, forms representations and predictions, and tests and refines those representations and predictions. It therefore stands to reason that the more opportunities one has for learning, the more one can in principle reduce one's uncertainty, assuming the environment has not changed in the meantime. Similarly, a reduction in the number of opportunities to learn will increase the uncertainty about whether one has accurately learned about the available behavioral options and their associated payoffs. We refer to the number of learning opportunities as an individual's effective lifespan to imply that it is the number of opportunities to learn throughout the individual's lifespan that matters here, not necessarily how long they live in the absence of relevant learning opportunities.

@@

We operationalized the four ``principal components'' of uncertainty that affect the
ability of agents to find and choose the optimal behavior. Recall these varieties of
uncertainty are \emph{environmental variability}; \emph{payoff ambiguity};
\emph{selection set size}; and the \emph{effective lifespan}. Environmental
variability is the probability that the behavior yielding the optimal payoff changes
from one generation to the next. We do not consider cases where the optimal behavior
varies within a generation. This is empirically justified since cultural evolution
occurs when meta-cognitive strategies and information are transferred via
learning and social influence, instead of by physical reproduction and death
as in genetic evolution.  The selection set size is the total number of possible
behaviors agents choose from, assumed to be constant within a simulation.  We
operationalized payoff ambiguity by manipulating the payoff difference between optimal
and non-optimal behaviors. The effective life span is the
number of time steps in which agents perform behaviors within a generation. 
Our model is not spatial, so agents have free, instant
access to all behaviors, and there is no spatial uncertainty, 
as might be the case for humans in a computer-based lab
experiment~\cite{McElreath2005,Morgan2012}.  Spatial dimensions could also be
considered irrelevant if costs associated with accessing a behavior were equal,
e.g., if it was assumed to be equally easy for bumblebees to fly to alternative
flower ``species''~\cite{Baracchi2018} or for great tits to fly to different food
sources~\cite{Aplin2017}.  Each simulated environment provides a distinct challenge
to the agents, which will lead agents to differently evolve to become social
learners or not, depending on which strategy optimizes net payoffs.

\emph{Agents} here are autonomous, intelligent, simulated problem solvers. Each
agent evaluates its environment and chooses how to act within that environment based
on a set of empirically-motivated rules.  Our agents select which behavior to do
probabilistically, weighted by the softmax function applied to the mean payoff they
have observed for each behavior. The softmax function is a biologically plausible
generalization of behavior selection under uncertainty that enables agents to often
greedily exploit the most lucrative behavior (based on their observations) but also
to sometimes explore alternatives~\cite{Schulz2019,Collins2013,Daw2006,Yechiam2005}.
Such individual-level intelligence is a critical adaptation that has apparently
co-evolved with the capacity for social learning, and has a marked influence on the
cultural evolution of social learning as we will demonstrate in our Analysis.
We will see that individual-level intelligence allows agents to recover from
the receipt of outdated information due to temporal environmental variability.

@@

The model environment provides behaviors for agents to perform. The structure
of the environment is specified by setting several parameters, including
three of the uncertainty parameters: environmental variability, selection set size, 
and payoff ambiguity. 
Each behavior is indexed by $b$ and modeled as a ``one-armed bandit'', which yields 
Bernoulli-distributed payoffs of 1 with probability $\pi_b$ and 0 with 
probability $1 - \pi_b$. $\pi_b$ is equivalently the expected
payoff of behavior/bandit $b$. The model environment affords $B$ behaviors, where
$B$ is the \emph{selection set size} uncertainty parameter, which
is constant throughout the course of a given simulation, but varied systematically
across simulations.

The model environment is dynamic: which behavior is optimal changes between
generations with probability equal to the \emph{environmental variability}, $u$.
The optimal behavior does not change within generations.
At the beginning of each generation, one bandit is optimal and yields expected
payoff $\pihigh$; the other $B-1$ behaviors yield a lower payoff, $\pilow$. 
The \emph{payoff ambiguity} here between optimal and non-optimal behavior
payoffs is defined by $\pihigh - \pilow$. We set $\pihigh = 0.9$ to be constant
across all simulations, so our analysis only varies payoff ambiguity by varying
$\pilow$. Specifically, payoff ambiguity increases with $\pilow$. 
By systematically varying these uncertainty parameters we change
the nature of the challenge faced by the agents. Each parameter individually 
can increase or decrease ``uncertainty,'' but it is not at all clear \emph{a priori}
how these uncertainties combine to affect the evolution of social learning.
These uncertainty parameters and the values we tested in our simulations
are summarized in Table~\ref{tab:uncertaintyParameters}.

@@

In this model, being a social learner is beneficial when it is more likely
on average to provide greater net payoffs than being an individual learner,
since evolution here acts as an optimization routine. 
Note that, unlike some models in the literature, it is not catastrophic for an 
individual to receive outdated information~\cite[e.g.]{Rogers1988}. Instead,
agents can explore alternatives and greedily perform a different observed optimal behavior
than an outdated optimal behavior they learned about socially. We rely
on this mechanism in the Analysis to explain the patterns we observe where
social learning evolves when $u > 0.5$, meaning that the
optimal behavior will switch between generations more often than not. If 
agents could not adjust to changing conditions like this, it seems like 
social learning would not evolve when $u > 0.5$.

@@

\paragraph{Intergenerational dynamics}
Between generations, agents first reproduce via asexual haploid reproduction; 
then new child agents learn from the
previous generation if they are social learners; and finally all agents from the
previous generation then die off (Figure~\ref{fig:schematic}C). 
$N$ reproducers are selected with replacement over $N$ independent draws, 
biased by performance:
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce in one draw}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
A child inherits its parent's social learning trait $s_i$ without mutation.
A social learner child with $s_i = 1$ learns from a teacher from its parent's
generation, including possibly their parent (oblique learning). 
A child selects its teacher by first selecting $N_T$ prospective
teachers from the population completely at random, then selecting the one with
highest net payoffs $\pi_i$ as their teacher, with ties broken randomly. While it
is somewhat arbitrary to first subset $N_T$, this is a conservative choice that
represnts the fact that access to the entire population is not generally guaranteed.
Social learners adopt their chosen teacher's ledger of expected payoffs $\ledger$,
and set all $c_{ib} = 1$ so expected payoffs remain within $[0, 1]$.  Asocial
learners ($s_i = 0$) are initialized with all $c_{ib} = 0$ and $\ledger = 0$ 
(Figure~\ref{fig:schematic}C). The
newly initialized population then begins behavior selection and payoff accumulation
for another $L$ time steps.

After reproduction and die-off, the optimal behavior changes with probability
$u$. If a random draw dictates the behavior should change, the new
behavior is chosen from all other behaviors with the current optimal behavior
excluded.

@@

Payoff ambiguity had a modest effect when $\pilow = 0.1$ increased to $\pilow = 0.45$,
causing social learning to be favored across more values of $u$, across all
selection set sizes and lifespans (first two rows of Figure~\ref{fig:mainResults}). 
This indicates that increased payoff ambiguity significantly increases the 
difficulty of discerning which behavior is optimal since agents evolve to 
risk social learning at greater values of $u$.
When $\pilow = 0.8$ the largest effect is smoothing and flattening of the $\meansl$ curves,
which indicates more drift (bottom row of Figure~\ref{fig:mainResults}). 
Initially this results in less frequent evolution
of social learning for the same values of $u$ and $B$. But after 
the $u$ value where $\meansl \approx 0.5$, social learning evolves more frequently than
for the same $u$, $B$, and $L$. When $\pilow \to \pihigh$, agents do better and better
whether they are social learners or not. We see that this leads to less penalty
for trying social information before possibly having to correct course due to
outdated social information.

@@

\mt{This subsec needs revision and associated figures need to be updated with 
$N=1000$ data}

@@

\mt{Need to make this section more specific in reviewing the findings.}
By systematizing four common uncertainty dimensions we developed a model of
the evolution of social learning that
predicts how different forms of uncertainty might interact to foster or suppress
social learning.  The model reproduced classic predictions that social learning can
yield higher payoffs than individual learning when uncertainty is sufficiently
limited.  By disambiguating and organizing various operationalizations of
uncertainty, we developed a more nuanced theoretical understanding of which forms
of uncertainty interact in which ways to make social learning more or less
beneficial compared to asocial learning.
Our model did not make any \emph{a priori} assumptions about whether it was
more costly for agents to be social or asocial learners, as some studies have
assumed or imposed~\cite{Rogers1988,Feldman1996,Morgan2012}. 
Instead, the uncertainty structure determined whether
or not social learning was beneficial or not, and evolution directed successive
generations towards the optimal strategy.
Such detailed theoretical models of uncertainty
are critical for understanding how humans adapt to an uncertain and
rapidly changing world, both to mitigate existential
threats~\cite{Moya2020,Jones2021}, and to capitalize on new opportunities such as
transitioning to clean energy use and
production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.

Our model necessarily made some simplifying assumptions that, while
empirically justified in many cases, may not always hold.
\mt{Review the ones mentioned in model section: non-spatial, no within-generation
environmental variability; no horizontal learning}.
We only considered
success-biased learning, although conformity is sometimes the dominant mechanism 
for social influence of behavior well~\cite{Muthukrishna2016a,Smaldino2018b}. 
Another implicit assumption
in our model was that the number of behaviors was constant, as was the relationship
between their payoffs---one was always optimal, and the rest were equally
suboptimal. However, this fails to account for evolutionary feedback which creates
new behavioral affordances as time progresses, e.g.\ via niche 
construction~\cite{Smaldino2012,Heras-Escribano2020}. Adjusting any of these
assumptions and running new computational analyses could yield further
understanding of the evolution of social learning across diverse contexts.

@@

We found that the evolution of social learning, under some modest
assumptions, is a complex phenomenon, dependent on several interrelated factors. By systematically
identifying common theoretical assumptions about uncertainty in the environments and contexts
in which social learning evolves, we were able to develop a more systematic understanding of
which uncertainty factors interact to determine whether social learning evolves. By
explicitly modeling a key factor of human intelligence, namely the ability to sense and
adjust to uncertainty and new information, we saw that social learning could be
valuable even if social learning provides outdated information. This work
informs most directly our current understanding of the evolution of social learning, but has
broader significance for the behavioral science of problem solving in uncertain conditions, and
for the design of artificial social intelligent systems in uncertain environments.  Our model
and its software implementation were designed to be modular and extensible.  Indeed,
we plan to extend this model to further develop a more detailed theoretical understanding of
the evolution of social learning; perhaps others will, too.
